MULTI EPOCH SPATIALLY RESOLVED RADIO OBSERVATIONS OF BETELGEUSE'S WIND ACCELERATION REGION

### 2014-07-14 draft1/ms.tex
Jul 14: onsala

		 2004 Oct 21,30  & 0.7 & $39\times 26$ & 0.37	& $99\pm 3$ & $92 \pm 3$& $92\pm 20$ &$28.68\pm 0.53$ & $2940\pm 170$\\
		 				 & 1.3 & $80\times 42$ & 0.09   & $127\pm 2$& $ 114\pm 2$ & $85\pm 5$& $13.83\pm 0.10$& $3140\pm 80$\\
						 & 2.0 & $121\times 91$ & 0.08  & $158\pm 6$& $ 152\pm 10$ & $132\pm 59$ & $7.23\pm 0.15$& $2270\pm 130$ \\
						 & 3.6 & $208\times 126$ & 0.02	& $215\pm 7$& $188\pm 7$ & $0\pm 10$& $3.34\pm 0.03$& $2110\pm 110$\\
						 & 6.1 & $377\times 264$ & 0.02	& $314\pm 30$& $203\pm 30$ & $0\pm 20$ & $1.55\pm 0.04$& $1140\pm 160$\\

@@

\rule{-2.6pt}{2.5ex}  2003 Aug 10,12 & 0.7 		& $40\times 27$ & 0.46	& $103\pm 4$& $92\pm 5$ & $104\pm 16$& $28.05\pm 0.84$& $2760\pm 230$ \\
									 & 1.3		& $80\times 42$ & 0.17	& $122\pm 5$& $114\pm 6$& $0\pm 5$& $11.20\pm 0.24$& $2490\pm 150$\\
									 & 2.0		& $119\times 96$& 0.10	& $132\pm 10$& $ 116\pm 9$& $11\pm 27$& $5.88\pm 0.17$&$3040\pm 360$\\
									 & 3.6		& $ 204\times 139$&	0.03& $193\pm 7$& $ 140\pm 11$ & $152\pm 7$& $2.80\pm 0.04$&  $2610\pm 170$\\
									 & 6.1 		& $378\times 297$&	0.03& $247\pm 47$& $179\pm 61$ & $169\pm 28$& $1.22\pm 0.04$&$2040\pm 680$ \\

@@

\rule{-2.6pt}{2.5ex}  2002 Apr 12,13  & 1.3 		& $91\times 59$ & 0.18	&$134\pm 9$ & $102\pm 6$& $36\pm 10$& $8.96\pm 0.24$& $2170\pm 250$\\
							& 2.0		&$131\times 98$ & 0.39	& $166\pm 16$& $104\pm 13$&$41\pm 11$ & $5.32\pm 0.24$& $2420\pm 450$ \\
							& 3.6		& $224\times 155$&0.03	& $234\pm 9$& $171\pm 10$& $41\pm7$& $2.66\pm 0.04$& $1690\pm 110$\\

@@

\rule{-2.6pt}{2.5ex}  2002 Feb 17,18 & 1.3 		& $83\times 48$&0.14	&$120\pm 4$ & $109 \pm 3$& $31\pm 13$&$10.87\pm 0.17$ & $2750\pm 140$\\
									& 2.0		& $128\times 90$&0.11	&$140\pm 13$& $135\pm 16$& $0\pm 28$&$5.38\pm 0.22$ & $2150\pm 300$\\
									& 3.6		& $200\times 135$&	0.03&$217\pm 9$ & $183\pm 10 $& $ 0\pm 5$ & $2.85\pm 0.04$& $1830\pm 110$\\

@@

\rule{-2.6pt}{2.5ex}  2001 Jan 02  & 1.3 		& $78\times 42$&0.08	&$124\pm 2$ & $115\pm 1$& $40\pm 8$& $12.58\pm0.08$& $2920\pm 70$ \\
 		 2000 Dec 23 & 0.7		& $44\times 20$& 0.18	&$98\pm 2$ &$91\pm 1$ &$0\pm 8$ & $29.02\pm 0.30$& $3070\pm 100$

@@

\subsection{Radio Diameters} 
\begin{figure}
\includegraphics[trim = 3mm 0mm 0mm 0mm, clip,height=12cm, width=8.5cm]{/home/eamon/pietown/paper/figs/fig2.ps}
\caption{Real component of the complex visibility plotted against the projected baseline length in wavelengths, $B_{\lambda}$, for the 2004 epoch at all wavelengths between $0.7-6.1$\,cm. The black data point contain only VLA antenna baselines, the red data points contain both VLA and PT antenna baselines, while the green data points contain only PT antenna baselines. The solid blue line represents the theoretical visibilities of a uniform intensity circular disk with an \textit{effective radius} defined as $\phi _{\rm{eff}}=0.5\sqrt{\theta _{\rm{maj}}\theta _{\rm{min}}}$.}
\label{fig2}
\end{figure}
It can be seen in Figure \ref{fig1} that the size of the source at 0.7\,cm is only a few beam diameters across. The source to beam ratio gets progressively smaller at longer wavelengths which makes fitting models to the images unreliable. In this case it is best to obtain estimates of the source size and flux density by fitting models directly to the calibrated complex visibilities. In Figure \ref{fig2} we plot the real component of the complex visibility against the projected baseline length in wavelengths, $B_{\lambda}$, for the 2004 epoch at all wavelengths between $0.7-6.1$\,cm. The data has been azimuthally vector averaged into uniformly sized bins over large portions of the unevenly sampled $u-v$ plane, where $u$ and $v$ are the spatial coordinates of the projected baseline in wavelengths, and $B_{\lambda} = \sqrt{u^2 + v^2}$. It is clear that the source is fully resolved at all wavelengths shown, i.e., we have obtained visibilities close to the first null. The VLA observations of \cite{lim_1998} only fully resolved Betelgeuse's atmosphere at 0.7\,cm at one epoch, but here we fully resolve the star at all wavelengths between 0.7 and 6.1\,cm at all our epochs. The black data points in Figure \ref{fig2} represent the real visibility measurements from the VLA antenna baselines only, the red data points include both VLA and PT antenna baselines, while the green data points include only PT antenna baselines. This highlights the importance of the inclusion of the PT antenna, especially at the longer wavelengths where it provides many baseline pairs close to the first null which helps to constrain the size of the extended atmosphere. The corresponding binned imaginary components of the complex visibility are not shown here but were mainly consistent with zero amplitude at all baselines, indicative of no significant asymmetrical structure.

Uniform intensity elliptical disk models were fitted to the complex visibilities as outlined in Section \ref{obs2.2}. The diameters and P.A.s of the best fits are listed in columns 5, 6, and 7 of Table \ref{tab1}. The source size is found to get progressively larger with increasing wavelength which implies that the radio surfaces are optically thick, in agreement with the findings of \cite{lim_1998}. At 0.7\,cm, these radio surfaces only slightly depart from spherical symmetry having axis ratios $\gtrsim 0.9$, in good agreement with the radio maps. This is also the case at 1.3 and 2.0\,cm except in 2002 April where the axis ratios change to $0.76\pm 0.07$ and $0.63\pm 0.10$ respectively, from near spherical symmetric values two months prior. At 3.5 and 6.1\,cm the radio surfaces display more consistent deviations from spherical symmetry having axis ratios ranging between $0.65\pm 0.10$ and $0.87\pm 0.04$. The intrinsic beam is elongated at all wavelengths due to inclusion of the PT antenna so caution must be exercised when reporting deviations from spherical symmetry.

In Figure \ref{fig2} we also plot the theoretical visibility curves of a uniform intensity circular disk with an \textit{effective radius} defined as $\phi _{\rm{eff}}=0.5\sqrt{\theta _{\rm{maj}}\theta _{\rm{min}}}$. The scatter between these model curves and the azimuthally averaged real visibility is low and indicates that circular disks represent the radio surfaces well. This is found to be the case at all epochs. Bearing in mind that the data is sparsely sampled between 2000 December and 2004 October we can cautiously investigate how the size of the atmosphere changes over this time frame. At 0.7 and 1.3\,cm there is no significant change in $\phi _{\rm{eff}}$ at all epochs with $\Delta \phi _{\rm{eff}} \lesssim 3\%$ about the mean, i.e., there is no significant change in the size of the inner atmosphere. We see more variability at longer wavelengths which sample the more extended atmosphere with $\Delta \phi _{\rm{eff}} \sim 14\%$ about the mean at 2.0 and 3.5\,cm. In fact at these two wavelengths, $\phi _{\rm{eff}}$ increases by $\sim 24\%$ in the 14 months between our 2003 and 2004 observations. We only have two measurements at 6.1\,cm and these again show an increase in $\phi _{\rm{eff}}$ between 2003 and 2004 with $\Delta \phi _{\rm{eff}} =20\%$, although the error are too large be be certain of this increase at 6.1\,cm.

\subsection{Flux Densities}
obtained from visibilty fits but also checked against elliptical Gaussian fitted results
plot flux versus frequency (see harper paris presentation)
$F_{\nu} = 0.185$\,mJy\,$\nu ^{1.33}$

\begin{figure}
\includegraphics[trim = 0mm 0mm 0mm 5mm, clip,scale=0.38, angle=90]{/home/eamon/pietown/paper/figs/fig3.ps}
\caption{xxx }
\label{fig3}
\end{figure}

\subsection{Temperature Profiles}
give more exact lim formula

\section{DISCUSSION}
\subsection{Radio Flux Density Variability} 
see Reid \& Menton 1996 conf proceedings\\
see Drake conf proceedings\\
e-MERLIN flux is concentrated\\
see Harper\_variability. ps\\
does flux go up as ang diam go up\\
see Harper 2001 discussion

>>>>

		 2004 Oct 21,30  & 0.7 & $39\times 26$ & 0.37	& $99\pm 3$ & $92 \pm 3$& $92\pm 20$ &$28.68\pm 0.53$ & $2940\pm 140$\\
		 				 & 1.3 & $80\times 42$ & 0.09   & $127\pm 2$& $ 114\pm 2$ & $85\pm 5$& $13.83\pm 0.10$& $3160\pm 80$\\
						 & 2.0 & $121\times 91$ & 0.08  & $158\pm 6$& $ 152\pm 10$ & $132\pm 59$ & $7.23\pm 0.15$& $2360\pm 190$ \\
						 & 3.6 & $208\times 126$ & 0.02	& $215\pm 7$& $188\pm 7$ & $0\pm 10$& $3.34\pm 0.03$& $2100\pm 110$\\
						 & 6.1 & $377\times 264$ & 0.02	& $314\pm 30$& $203\pm 30$ & $0\pm 20$ & $1.55\pm 0.04$& $1770\pm 320$\\

@@

\rule{-2.6pt}{2.5ex}  2003 Aug 10,12 & 0.7 		& $40\times 27$ & 0.46	& $103\pm 4$& $92\pm 5$ & $104\pm 16$& $28.05\pm 0.84$& $2760\pm 200$ \\
									 & 1.3		& $80\times 42$ & 0.17	& $122\pm 5$& $114\pm 6$& $0\pm 5$& $11.20\pm 0.24$& $2670\pm 190$\\
									 & 2.0		& $119\times 96$& 0.10	& $132\pm 10$& $ 116\pm 9$& $11\pm 27$& $5.88\pm 0.17$&$3010\pm 340$\\
									 & 3.6		& $ 204\times 139$&	0.03& $193\pm 7$& $ 140\pm 11$ & $152\pm 7$& $2.80\pm 0.04$&  $2630\pm 230$\\
									 & 6.1 		& $378\times 297$&	0.03& $247\pm 47$& $179\pm 61$ & $169\pm 28$& $1.22\pm 0.04$&$2010\pm 780$ \\

@@

\rule{-2.6pt}{2.5ex}  2002 Apr 12,13  & 1.3 		& $91\times 59$ & 0.18	&$134\pm 9$ & $102\pm 6$& $36\pm 10$& $8.96\pm 0.24$& $2170\pm 200$\\
							& 2.0		&$131\times 98$ & 0.39	& $166\pm 16$& $104\pm 13$&$41\pm 11$ & $5.32\pm 0.24$& $2420\pm 400$ \\
							& 3.6		& $224\times 155$&0.03	& $234\pm 9$& $171\pm 10$& $41\pm7$& $2.66\pm 0.04$& $1690\pm 120$\\

@@

\rule{-2.6pt}{2.5ex}  2002 Feb 17,18 & 1.3 		& $83\times 48$&0.14	&$120\pm 4$ & $109 \pm 3$& $31\pm 13$&$10.87\pm 0.17$ & $2750\pm 130$\\
									& 2.0		& $128\times 90$&0.11	&$140\pm 13$& $135\pm 16$& $0\pm 28$&$5.38\pm 0.22$ & $2230\pm 350$\\
									& 3.6		& $200\times 135$&	0.03&$217\pm 9$ & $183\pm 10 $& $ 0\pm 5$ & $2.85\pm 0.04$& $1820\pm 130$\\

@@

\rule{-2.6pt}{2.5ex}  2001 Jan 02  & 1.3 		& $78\times 42$&0.08	&$124\pm 2$ & $115\pm 1$& $40\pm 8$& $12.58\pm0.08$& $2920\pm 60$ \\
 		 2000 Dec 23 & 0.7		& $44\times 20$& 0.18	&$98\pm 2$ &$91\pm 1$ &$0\pm 8$ & $29.02\pm 0.30$& $3040\pm 80$

@@

\subsection{Radio Diameters}

\begin{figure}
\includegraphics[trim = 3mm 0mm 0mm 0mm, clip,height=12cm, width=8.5cm]{/home/eamon/pietown/paper/figs/fig2.ps}
\caption{Real component of the complex visibility plotted against the projected baseline length in wavelengths, $B_{\lambda}$, for the 2004 epoch at all wavelengths between $0.7-6.1$\,cm. The black data point contain only VLA antenna baselines, the red data points contain both VLA and PT antenna baselines, while the green data points contain only PT antenna baselines. The solid blue line represents the theoretical visibilities of a uniform intensity circular disk with an \textit{effective radius} defined as $\phi _{\rm{eff}}=0.5\sqrt{\theta _{\rm{maj}}\theta _{\rm{min}}}$.}
\label{fig2}
\end{figure}

It can be seen in Figure \ref{fig1} that the size of the source at 0.7\,cm is only a few beam diameters across. The source to beam ratio gets progressively smaller at longer wavelengths which makes fitting models to the images unreliable. In this case it is best to obtain estimates of the source size and flux density by fitting models directly to the calibrated complex visibilities. In Figure \ref{fig2} we plot the real component of the complex visibility against the projected baseline length in wavelengths, $B_{\lambda}$, for the 2004 epoch at all wavelengths between $0.7-6.1$\,cm. The data has been azimuthally vector averaged into uniformly sized bins over large portions of the unevenly sampled $u-v$ plane, where $u$ and $v$ are the spatial coordinates of the projected baseline in wavelengths, and $B_{\lambda} = \sqrt{u^2 + v^2}$. It is clear that the source is fully resolved at all wavelengths shown, i.e., we have obtained visibilities close to the first null. The VLA observations of \cite{lim_1998} only fully resolved Betelgeuse's atmosphere at 0.7\,cm at one epoch, but here we fully resolve the atmosphere at all wavelengths between 0.7 and 6.1\,cm at all our epochs. The black data points in Figure \ref{fig2} represent the real visibility measurements from the VLA antenna baselines only, the red data points include both VLA and PT antenna baselines, while the green data points include only PT antenna baselines. This highlights the importance of the inclusion of the PT antenna, especially at the longer wavelengths where it provides many baseline pairs close to the first null which helps to constrain the size of the extended atmosphere. The corresponding binned imaginary components of the complex visibility are not shown here but were mainly consistent with zero amplitude at all baselines, indicative of no significant asymmetrical structure.

Uniform intensity elliptical disk models were fitted to the complex visibilities as outlined in Section \ref{obs2.2}. The diameters and P.A.s of the best fits are listed in columns 5, 6, and 7 of Table \ref{tab1}. The source size is found to get progressively larger with increasing wavelength which implies that the radio surfaces are optically thick, in agreement with the findings of \cite{lim_1998}. At 0.7\,cm, these radio surfaces only slightly depart from spherical symmetry having axis ratios $\gtrsim 0.9$, in good agreement with the radio maps. This is also the case at 1.3 and 2.0\,cm except in 2002 April where the axis ratios change to $0.76\pm 0.07$ and $0.63\pm 0.10$ respectively, from near spherical symmetric values two months prior. At 3.5 and 6.1\,cm the radio surfaces display more consistent deviations from spherical symmetry having axis ratios ranging between $0.65\pm 0.10$ and $0.87\pm 0.04$.

The intrinsic beam is elongated at all wavelengths due to inclusion of the PT antenna so caution must be exercised when reporting deviations from spherical symmetry. In Figure \ref{fig2} we include the theoretical visibility curves of a uniform intensity circular disk with an \textit{effective radius} defined as $\phi _{\rm{eff}}=0.5\sqrt{\theta _{\rm{maj}}\theta _{\rm{min}}}$. The scatter between these model curves and the azimuthally averaged real visibility is low and indicates that circular disks represent the radio surfaces well. This is found to be the case at all epochs. Bearing in mind that the data is sparsely sampled between 2000 December and 2004 October we can cautiously investigate how the size of the atmosphere changes over this time frame. At 0.7 and 1.3\,cm there is no significant change in $\phi _{\rm{eff}}$ at all epochs with $\Delta \phi _{\rm{eff}} \lesssim 3\%$ about the mean, i.e., there is no significant change in the size of the inner atmosphere. We see more variability at longer wavelengths which sample the more extended atmosphere with $\Delta \phi _{\rm{eff}} \sim 14\%$ about the mean at 2.0 and 3.5\,cm. In fact at these two wavelengths, $\phi _{\rm{eff}}$ increases by $\sim 24\%$ in the 14 months between our 2003 and 2004 observations. We only have two measurements at 6.1\,cm and these again show an increase in $\phi _{\rm{eff}}$ between 2003 and 2004 with $\Delta \phi _{\rm{eff}} =20\%$, although the error are too large be be certain of this increase at 6.1\,cm.

\subsection{Flux Densities}
\begin{figure}
\includegraphics[trim = 0mm 0mm 0mm 5mm, clip,scale=0.38, angle=90]{/home/eamon/pietown/paper/figs/fig3.ps}
\caption{Radio spectra for Betelgeuse at VLA wavelengths during $1972 - 2004$. The continuous line represents the radio spectrum from \cite{newell_1982} while the dashed line is the almost identical spectrum from  \cite{drake_1992}. The dash-dotted line is the radio spectrum that best fits the more recent data between $1996 - 2004$.}
\label{fig3}
\end{figure}

Betelgeuse's atmosphere is optically thick at radio wavelengths which allows its radio flux density to be characterized as resulting from an opaque disk with $F_{\nu} \propto T_{e}\phi _{\rm{eff}}^2$, where $T_{e}$ is the electron temperature. The relationship tells us that radio flux density variations are caused by either changes in the radio disk size or gas temperature distribution. The size of the radio emitting region is controlled mainly by the distribution of ionized material. \cite{newell_1982} found that multi-wavelength radio flux density measurements from the period $1972-1981$ could be described by the radio spectrum $0.24\nu ^{1.32}_{\rm{GHz}}$, while \cite{drake_1992} reported an almost identical spectrum for data spanning $1986-1990$.  In Figure \ref{fig3} we plot these radio spectra along with our VLA/PT flux density values and the values from \cite{lim_1998}. It can be seen that all the data from 1996 onwards lie below the spectrum of \cite{newell_1982}. We find that these observations can be better fitted by a spectrum described by $F_{\nu} = 0.185$\,mJy\,$\nu ^{1.33}$. This indicates that there has been an overall reduction in the radio flux over the last few decades which would mean that either the global temperature in the wind acceleration has decreased or that its size has decreased.

\cite{bookbinder_1987} found that the 6\,cm flux can vary by as much as $\sim 40\%$ over a 20 day timescale, while \cite{drake_1992} monitored the star over a four year period and reported stochastic flux variability at 2.0 and 3.6\,cm of $\lesssim 25\%$ and possibly at 6\,cm on a timescale $\lesssim 1$ month. Here we report on how the radio flux has varied during the period $1996-2004$ using the measurements of \cite{lim_1998} and the VLA-Pie Town data.

We plot the resulting flux densities against time in Figure \ref{fig:5.16} and also include the measurements of \cite{lim_1998}. We find significant variability at all wavelengths and at all epochs apart from the 2002 data sets which were only obtained two months apart from each other. Over the 10\,yr period, the maximum variability is found to be 23, 27, 32, 21, and $35\%$ at 0.7, 1.3, 2.0, 3.5, and 6.2\,cm respectively, which are similar to those reported by \cite{drake_1992} and \cite{bookbinder_1987}. The shortest wavelengths (i.e., 0.7 and 1.3\,cm) are the most sampled and although there are only five data points at these wavelengths to compare, there does appear to be a correlation in their flux variability.

However, the flux variability may be a manifestation of significant structural changes or thermal changes over the order of only a few years.  This adds weight to the idea discussed in Section \ref{sec:5.13} that the e-MERLIN feature(s) are optically thin at the shorter VLA wavelengths and therefore are not detectable in our highest resolution data sets. Nevertheless, the large variations at 6\,cm could still indicate that these features undergo large structural changes on timescales of only a few years or perhaps even shorter.

\subsection{Temperature Profiles}
Fitting circular disks to the measured visibilities was a method used by \cite{lim_1998} to quantify not only the total flux, but also the area of the region where the radio emission is emanating from. Knowing this area, the brightness temperature (and thus the electron temperature) can be calculated using Equation \ref{eq:1.43aa}. We followed the same method and applied this equation to our VLA-Pie Town data. The derived electron temperature as a function of distance from the star is plotted in Figure \ref{fig:5.16a} along with the values from \cite{lim_1998} and the single value from \cite{richards_2013} (which was derived by fitting a 2-D ellipse to the low resolution e-MERLIN data set and finding the area of the Gaussian). Although there appears to be some variability in the mean electron temperature at various epochs, our values are consistent with a relatively low temperature wind, which was one of the main findings of \cite{lim_1998}.

A detailed discussion on how spectral indices can be used to infer properties of stellar winds is presented in Chapter \ref{chap:6}. For now, we note that if the electron temperature and density are represented as different power laws in radius, then a simple relationship  exists between the power-law coefficients and the spectral index (see section \ref{sec:6.6}). The mean spectral index between 4.5 and 43\,GHz over all epochs is found to be $\alpha =1.31$, while the dashed line in Figure \ref{fig:5.16a} represents the derived power-law fit, $T_{e} \propto r^{-0.79}$, to the electron temperature. These two values can then be used to derive a power law for the density of $n_{e} \propto r^{-2.93}$, which is in excellent agreement with the semi-empirically derived value of $n_{e} \propto r^{-2.85}$ from \cite{harper_2001}. Compare to ogorman paper?

give more exact lim formula

\begin{figure}
\includegraphics[trim = 0mm 0mm 0mm 5mm, clip,scale=0.38, angle=90]{/home/eamon/pietown/paper/figs/fig4.ps}
\caption{xxx }
\label{fig3}
\end{figure}

\section{DISCUSSION}
\subsection{Radio Flux Density Variability} 
see Reid \& Menton 1996 conf proceedings\\
see Drake conf proceedings\\
e-MERLIN flux is concentrated\\
see Harper\_variability. ps\\
does flux go up as ang diam go up\\
see Harper 2001 discussion
The dashed red line represents the 6\,cm flux of \cite{richards_2013} which lies within the range of 6\,cm values observed between 1996 and 2004. This is surprising because most of the 6\,cm flux in the e-MERLIN map is contained in the two bright features and, as we see no evidence for these features in our VLA-Pie Town data, we would expect the 6\,cm flux in the e-MERLIN map to be larger.
